---
title: "GLM of count data"
author: "J skelton"
date: "`r Sys.Date()`"
output: html_document
---


# Cray cray worm count example

## Getting started

* Set your WD
* Load the data.frame of crayfish and their worms (a copy is on Bb)
* install MASS and lmtest packages if you dont have them

```{r}
# install.packages("MASS")
# install.packages("lmtest")
#install.packages("AER")

library(MASS)
library(lmtest)
library(AER)
#read in raw data
df <- read.csv(file = "cray_worm_NewRiver(1).csv")
```

* Make two new variables, one for total number of worms on each crayfish and one logical variable for whether the crayfish is native (true) or not (false).

* Then subset the df to include the variables native, sex, CL, mass, and number of worms.

```{r}
#make a new variable of the sum of all worm species for each crayfish
df$worms <- rowSums(df[,14:23])

#make a true/false variable for whether 
#it is a native crayfish species
#the native species are coded "bart" and "scio" in the data
df$native <- df$SPEC %in% c("scio","bart")

#subset only the variables of interest
df <- df[,c("native","SEX","CL","MASS", "worms")]

#remove observations with missing data
df <- na.omit(df)
```

# Modeling

To save time, we'll skip model selection procedure, since we've already identified the most important predictors during our multiple regression example last week.

We'll start with looking at the data

```{r, fig.width = 12}
plot(df)
```

Clearly CL and MASS are strongly correlated. We'll keep CL as our measure for size, and Sex and native as other predictors of worm abundance.

Plot CL versus worms. Does this look linear? Does it satisfy the assumptuon of homogeniety? Other issues?

```{r}
plot(df$worms ~ df$CL, ylab = "number of worms", xlab = "carapace length (mm)")
```

Becasue we are dealing with counts, we'll start with a Poisson model.

```{r}
mod.pois <- glm(worms ~
                  CL  + native, 
                data = df, family = "poisson")
```

## Notice:
* We don't need to specify the link bc a log link is assumed for poisson GLM

Look at the model summary
```{r}
summary(mod.pois)
```

* Which predictors are significant?
* What is our R square?

# Model validation

Is there evidence of over-dispersion?
* One way to check is to divide the residual deviance by degrees of freedom. If the quotient is much greater than 1, we need to deal with over-dispersion.

* 23022/620 = 37.13, so yeah, looks like we got a problem.

Here is another way. Look at the residuals versus the predictor CL.
**note that I have NOT used mod.pois$residuals. Why?**

```{r}
plot(predict(mod.pois),residuals.glm(mod.pois, type = "deviance"))
```

We still have a huge trumpet - so, again, more confirmation that we have overdispersion in our response variable. 

In case you need further convincing, yes, there are statistical tests. The one implemented here tests the null hypothesis that the mean and variance are equal (i.e. n over dispersion).

```{r}
dispersiontest(mod.pois)
```
p val says dist. does not fit data very well

# Negative binomial model

Now we will try a negative binomial GLM which has an extra parameter to estimate the relationship between the mean and variance of our model predictions.

```{r}
mod.nb <- glm.nb(worms ~
                   CL + native, 
                 data = df)

summary(mod.nb)
```

* Now look at residual deviance vs df

* Look at theta. What is that? 

# Compare residuals from the 2 models

```{r}
#first poisson
#predict.glm defaults to standardized deviance residuals
par(mfrow = c(1,2))
plot(predict(mod.pois),residuals.glm(mod.pois), main = "poisson glm")
#ideally deviance residuals will fall between -2 and 2
#we'll draw some ref lines
abline(h = c(2,-2), col = "red")

#now the NB model
plot(predict(mod.nb),residuals.glm(mod.nb), main = "NB glm")
abline(h = c(2,-2), col = "red")
```

# Directly compare models

Clearly the residuals look much better for the negative binomial model.
Still unconvinced? Because Poisson is a special case of neg binomial where theta (k) = infinity, these are nest models and can be directly compared using a Liklihood ratio test.

```{r}
lrtest(mod.pois,mod.nb)
```

Log liklihood is significantly higher for model 2 - the negative binomial.

* Conclusion, better fit with NB :D

# Now lets plot the model

```{r}
par(mfrow = c(1,1))

new.dat2 <- expand.grid(CL = 5:50, native = c(TRUE,FALSE), SEX ="M1")
pred2 <- predict(mod.nb, newdata = new.dat2, se.fit = T, type = "response")
#add newly random generated data (newdat2), fit SE to later make CI bars, 
#type = response EXPONENTIATES the log linked predictions
#necessary bc our model predicts the log number of worms

plot(1,1, type = "n",
     ylim = c(0,350), xlim = c(5,50),
     ylab = "number of worms", xlab = "carapace length (mm)")


points(df$CL[df$native == "FALSE"], 
       df$worms[df$native == "FALSE"],
       pch =21, bg = rgb(134/255, 31/255, 65/255, .25), cex = 1)
points(df$CL[df$native == "TRUE"], 
       df$worms[df$native == "TRUE"],
       pch =21, bg = rgb(232/255, 119/255, 34/255, 1), cex = 1)



lines(new.dat2$CL[new.dat2$native == "TRUE"], pred2$fit[new.dat2$native == "TRUE"], 
      col = rgb(232/255, 119/255, 34/255, 1), lwd = 2)
lines(new.dat2$CL[new.dat2$native == "FALSE"], pred2$fit[new.dat2$native == "FALSE"], 
      col = rgb(134/255, 31/255, 65/255, 1), lwd = 2)

legend(5,370, pch = 21, 
       pt.bg = c(rgb(232/255, 119/255, 34/255,1), rgb(134/255, 31/255, 65/255,.25)),
       pt.cex = 1.5, pt.lwd = 2,    legend = c("non-native host","native host"), bty = "n")



lwr.inv <- pred2$fit[new.dat2$native == "TRUE"] - pred2$se.fit[new.dat2$native == "TRUE"]
upr.inv <- pred2$fit[new.dat2$native == "TRUE"] + pred2$se.fit[new.dat2$native == "TRUE"]
lwr.non <- pred2$fit[new.dat2$native == "FALSE"] - pred2$se.fit[new.dat2$native == "FALSE"]
upr.non <- pred2$fit[new.dat2$native == "FALSE"] + pred2$se.fit[new.dat2$native == "FALSE"]

polygon(x = c(5:50,rev(5:50)), 
        y = c(lwr.inv,rev(upr.inv)), 
        col = rgb(232/255, 119/255, 34/255, 0.25), border = F)

polygon(x = c(5:50,rev(5:50)), 
        y = c(lwr.non,rev(upr.non)), 
        col = rgb(134/255, 31/255, 65/255, 0.25), border = F)

```






