---
title: "Lab 5"
author: "Kent Codding"
date: "2023-10-05"
output: word_document
---

```{r}
library(MASS)
library(lmtest)
setwd("C:/Users/Kent Codding/Desktop/Adv Biostat/Lab 5")
df <- read.csv("BACS.csv", row.names = 1)
```

# Question 1
```{r}
plot(df[-1]) #plots all potential explanatory variables (not including response var - BACS)
```
no obvious collinearity, but considering biologicical phenomena, it would make sense that hardwood cover percentage increases with years after burn. This also may confound with Burn_Season.
```{r}
# use pearson's r to check
cor(df$hardwood_cov_perc, df$years_since_burn) #not collinear
# cor(df$hardwood_cov_perc, df$Burn_Season) not numeric


```
My hypothesis was incorrect. The hardwood cover percentage does not correlate enough with years after burn for it to be an issue
# Question 2
justification for error distribution: since we are looking at count data, which is a poisson process, I will start with a poisson distribution to determine whether the residual deviance displays overdispersion.

```{r}
mod1p <- glm(BACS ~ ., data = df, family = "poisson") 
summary(mod1p)
```
```{r}
mod1p$deviance/mod1p$df.residual > 1.5 #if TRUE, then overdispersion is apparent
```
Yes, there is overdispersion :( now try negative binomial.
```{r}
mod1nb <- glm.nb(BACS ~ ., data = df) 
summary(mod1nb)
```

```{r}
mod1nb$deviance / mod1nb$df.residual
lrtest(mod1p,mod1nb)

```
The residual deviance divided by degrees of freedom is much closer to 1. Now plot the residuals for each model. Log likelihood is also higher for the negative binomial model

```{r}
#poisson
par(mfrow = c(1,2))
plot(predict(mod1p),residuals.glm(mod1p), 
     main = "poisson glm",
     xlab = "predicted values",
     ylab = "standardized deviance residuals")
#ideally deviance residuals will fall between -2 and 2
abline(h = c(2,-2), col = "coral")

#now the NB model
plot(predict(mod1nb),residuals.glm(mod1nb), main = "NB glm",
     xlab = "predicted values",
     ylab = "standardized deviance residuals")
abline(h = c(2,-2), col = "coral")
```
Now that I have explored the residual distributions of the two models, I have selected the negative binomial model as it performed better on each of the three tests: deviance/degrees of freedom, log likelihood, and deviance falling between (mostly) +/- 2 for predicted values.

# Question 3
```{r}
#drop 1 method
drop1(mod1nb, test = "F")
```

```{r}
mod2 <- glm.nb(BACS ~ area_large_pines + years_since_burn + Burn_Season + 
    Stand_age + hardwood_cov_perc, data = df)
summary(mod2)
```
```{r}
drop1(mod2, test = "F")
```
```{r}
mod3 <- glm.nb(BACS ~ years_since_burn +      Burn_Season + 
    Stand_age + 
    hardwood_cov_perc,
    data = df)
summary(mod3)
```

# Question 4

calculate pseudo R-squared for GLM
```{r}
# residual deviance/ null deviance
mod3$deviance / mod3$null.deviance
```
The final model is useful as the R^2 value is above 0.4, which is typically "good enough for ecology" even though the model is not perfect.

Based on the z-scores, the two most significant predictors are hardwood_cover_perc and years_since_burn in that order. Moreover, both are negatively correlated, suggesting that Bachman's sparrows dislike luxurious, unburned forests with a high hardwood tree density and like less dense and more recently burned forests.

# Question 5 - Crayfish!
read in VT crayfish data
```{r}
cray <- read.csv("cray_groom.csv", row.names = 1)
```

visualize data
```{r}
plot(cray)
```

It looks like experimental design purposefully avoided collinearity among predictors. CL appears to correlate most with grooming based on this visualization.

```{r}
plot(cray$groomed ~ cray$cl) 
```
Evidently, the response variable only has two levels, so a binomial error distribution is most appropriate.
```{r}
cray.mod1 <- glm(groomed ~ sex + cl + worm_spp,
                 data = cray,
                 family = "binomial")
summary(cray.mod1)
```
```{r}
drop1(cray.mod1, test = "Chisq")
```
worm_spp has the least significant effect based on the Chi-squared test and the AIC, so worm_spp should be removed. This will remove the 3 levels displayed in the previous summary.

```{r}
cray.mod2 <- glm(groomed ~ sex + cl,
                 data = cray,
                 family = "binomial")
summary(cray.mod2)
```
```{r}
drop1(cray.mod2, test = "Chisq")

```
sex has an insignificant effect based on the Chi-squared test, Likelihood ratio test, and the AIC, so sex should be removed.

## Final Model
```{r}
ultimate.cray.mod <- glm(groomed ~ cl,
     data = cray,
     family = "binomial")
summary(ultimate.cray.mod)
```

## plot predictor and grooming response
```{r}
y_pred <- predict(ultimate.cray.mod, type = "response")
plot(y_pred ~ cray$cl,
     main = "Predicted grooming based on carapace length",
     ylab = "Probability of Grooming Response",
     pch = 21,
     bg = "turquoise")
```

## for fun 
I hypothesize that larger crayfish do not groom as much due to the previous relationship that we observed between carapace length and log number of worms. For instance, a larger crayfish is already expected to be free real estate for worms, so natural selection likely selected for larger crayfish that decided to gamble and let their body become a haven for worms of all kinds, *sometimes* living longer lives depending on if the large crayfish gets a royal flush with all mutualistic worms or a 2 and 7 off-suit with all parasitic worms.

## more fun 
```{r}
plot(cray$groomed ~ cray$cl,
     main = "Observed vs Predicted Grooming Based on Carapace Length",
     ylab = "Probability of Grooming Response",
     pch = 21,
     bg = "firebrick1")
lines(y_pred ~ cray$cl,
      col = "turquoise")
legend(35, 1, c("observed", "predicted"),
       pch = 21,
       pt.bg = c("firebrick1", "turquoise"))
```

