---
title: "Lab 2"
author: "Kent Codding"
date: "2023-09-07"
output: word_document
---
# Read in Data
```{r}
setwd("C:/Users/Kent Codding/Desktop/Adv Biostat/Lab 2") #set wd to source file location
case.data <- read.csv("united_states_covid19_hospitalizations_deaths_and_ed_visits_by_state.csv", skip = 2, row.names = 1)
vax.data <- read.csv("covid19_vaccinations_in_the_united_states.csv" , skip = 5, row.names = 1)
```
# Operation 1
```{r}
head(case.data)
```
# keep certain names
```{r}
names(case.data) 
case.data <- case.data[,c("Total.Deaths", "Death.Rate.per.100000.for.the.Past.3.Months", "Total.Death.rate.per.100000", "New.COVID.19.hospital.admissions.per.100000.population.past.week..total.")] 
head(case.data)
```

# Operation 2
```{r}
names(vax.data) 
vax.data <- vax.data[,c("Percent.of.total.pop.with.at.least.one.dose", "Percent.of.total.pop.with.a.completed.primary.series")] 
head(vax.data)
```
# Merge the data
```{r}
state.name[!(state.name %in% rownames(case.data))]

```
# Operation 3
```{r}
state.name[!(state.name %in% rownames(vax.data))]
rownames(vax.data)
```

# Question 1
  at index 44 in the output from rownames(vax.data) above, New York is listed as "New York State." Thus, this line of code -  state.name[!(state.name %in% rownames(vax.data))]
 - that checks for state names where the state name Character Object is NOT in the rownames of vax data will return "New York" because R reads "New York State" as a different character object than "New York."
 
# Question 4
```{r}
rownames(vax.data)[44] <- "New York"
row.names(vax.data)
```

# Operation 5
```{r}
all.data <- merge(vax.data, case.data, by = 0) #merges vax.data and case.data by row

```

# Operation 6
```{r}
all.data <- data.frame(all.data, row.names = all.data$Row.names) #removes numerical row names (indices) and replaces with state names 
all.data <- all.data[,-1] #gets rid of first column that contained redundant row names
rownames(all.data)
```

# Operation 7
```{r}
library(magrittr) # load magrittr for pipe operator
subset(all.data, row.names(all.data) %in% state.name) %>% row.names() #check if this subset command only includes 50 states as rownames
all.data <- subset(all.data, row.names(all.data) %in% state.name) #get rid of territories
names(all.data) <- c("perc_one_dose","perc_fully_vax",
                     "total_deaths","death_rate_3_months",
                     "total_death_rate","new_case_rate_this_week") #make better column names
```
# Operation 8
```{r}
str(all.data)
all.data <- data.frame(lapply(all.data, as.numeric),
                       row.names = rownames(all.data))
```

# Question 2
  R gave us the warning message "NAs introduced by coercion" because some character values like "Data not available" found in all.data cannot be converted to a number. Instead, R coerces these values to NAs.
  
# Read in usdata
```{r}
library(usdata)
data(state_stats)
state_stats <- data.frame(state_stats, row.names = 1)
```
# Operation 9
```{r}
all.data <- merge(all.data, state_stats, by = 0) #merges vax.data and case.data by row
head(all.data)
```
## write to csv
```{r}
write.csv(all.data, file = "all_data.csv")
```

